---
documentclass: nojss
author:
  - name: Alex Asher
    affiliation: Texas A\&M University
    address: >
      Department of Statistics
    email: \email{alexasher@stat.tamu.edu}
title:
  formatted: "Semiparametric Analysis of Polygenic Gene-Environment Interactions in Case-Control Studies with \\pkg{caseControlGE}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Semiparametric Analysis of Polygenic Gene-Environment Interactions in Case-Control Studies with caseControlGE"
  # For running headers, if needed
  short:     "\\pkg{caseControlGE}: Gene-Environment Interactions in Case-Control Studies"
abstract: >
  Gene-environment interactions can be efficiently estimated in case-control data by methods that assume gene-environment independence in the source population, but until recently such techniques required parametric modelling of the genetic variables.  The \pkg{caseControlGE} package implements the methods of \citeauthor{Stalder2017} (2017, \emph{Biometrika}, \textbf{104}, 801-812) and \citeauthor{Wang2018unpublished} (2018, unpublished), which exploit the assumption of gene-environment independence without placing any assumptions on the marginal distributions of the genetic or environmental variables.  These methods are ideally suited for analyzing complex polygenic data for which parametric distributional models are not feasible.  In addition to the two estimators, the package also supplies a function to simulate case-control data and several helper functions for use on model objects.  Use of this package is illustrated by simulating and analyzing data from a case-control study of breast cancer.
keywords:
  # at least one keyword must be supplied
  formatted: [case-control study; gene-environment interaction; genetic epidemiology; retrospective method; semiparametric analysis; polygenic analysis]
  plain: [case-control study; gene-environment interaction; genetic epidemiology; retrospective method; semiparametric analysis; polygenic analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  <!-- \usepackage{pdfpages} -->
  <!-- \usepackage[letterpaper]{geometry} -->
output: rticles::jss_article
vignette: >
  %\VignetteIndexEntry{Semiparametric Analysis of Polygenic Gene-Environment Interactions in Case-Control Studies with caseControlGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\begin{titlepage}
\end{titlepage}

\def\bbeta{\mbox{\boldmath $\beta$}}
\def\pr{\hbox{pr}}
\def\E{{\mathbf E}}

```{r setup, cache=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE, cache=TRUE, autodep=TRUE
  # , echo=TRUE
  # , comment = "#>"
)
```

\section{Introduction}

\subsection{caseControlGE package}

The \pkg{caseControlGE} package \citep{Asher2018R} contains tools for the analysis of case-control data using \proglang{R} \citep{R2018}.  It implements the methods of \citet{Stalder2017} and \citet{Wang2018unpublished}, both of which fall under the class of semiparametric retrospective profile likelihood estimators.  These methods are the first available to exploit the assumption of gene-environment independence while treating the genetic component nonparametrically.  As such, they are well suited to replace logistic regression as the preferred method in situations where parametric distributional models are not feasible, such as in the analysis of complex polygenic data.

\pkg{caseControlGE} contains three main functions: \code{simulateCC}, \code{spmle}, and \code{spmleCombo}, as well as several helper functions.  Section \ref{sec:simulateCC} of this paper introduces \code{simulateCC} in the context of simulating case-control data analogous to the data analyzed in \cite{Wang2018unpublished}.  Section \ref{sec:spmle} introduces \code{spmle} as a tool to analyze the simulated data, and section \ref{sec:spmleCombo} introduces \code{spmleCombo} to conduct a more efficient analysis of the simulated data.

\subsection{Background}

Case-control studies are retrospective observational studies in which the sample consists of a group of healthy subjects and a group of diseased subjects.  A crucial aspect of the case-control design is that the outcome, disease status, is known \emph{before} sampling.  The ability to deliberately oversample diseased subjects makes the case-control design cost effective, which is why it is widely popular in studies of gene-environment interactions.

Given the genetic and environmental covariates $G$ and $E$, we assume the risk of disease $D$ in the underlying population follows the model \[ \pr(D=1 \mid G,X) = H\{\alpha_0 + m(G,X,\bbeta)\}, \] where $H(x)=\{ 1 + \exp(-x)\}^{-1}$ is the logistic distribution function and $m(G,X,\bbeta)$ is a function that describes the joint effect of $G$ and $X$ and is known up to the unspecified parameters of interest $\bbeta$.  

Given the retrospective nature of case-control sampling, it is surprising that standard prospective logistic regression can be used to obtain unbiased estimates of $\bbeta$ \citep{PrenticePyke1979}.  Logistic regression requires no assumptions about the joint distribution of $G$ and $E$, but it suffers from low power when estimating $G * E$ interaction effects.  To gain efficiency, \citet{ChatterjeeCarroll2005} exploited the assumption of gene-environment independence in the source population to maximize the retrospective likelihood while profiling out the distribution of $E$.  Their method is available as the function \code{snp.logistic} in the \emph{Bioconductor} package \pkg{CGEN} \citep{CGEN2012}.

The method of \citeauthor{ChatterjeeCarroll2005}, and subsequent methods utilizing the same retrospective profile likelihood framework, require a parametric model for the distribution of $G$ given $E$.  This becomes difficult as the number and complexity of genetic variables in the model grows.  Capitalizing on advances in high-throughput genomics, genome-wide association studies have identified scores of SNPs associated with complex diseases such as cancers and diabetes.  Modern case-control studies of gene-environment interactions need efficient methodology that allows for a flexible and arbitrarily complex genetic component, such as multiple correlated SNPs and/or continuous polygenic risk scores.

The method of \citet{Stalder2017} extends the retrospective profile likelihood framework of \citeauthor{ChatterjeeCarroll2005}, dispensing with the need to model $G$ parametrically.  When the population disease rate $\pi_1$ is known, the retrospective profile loglikelihood can be estimated (up to an additive constant) using just the case-control sample and without modeling the distribution of $G$.  When $\pi_1$ is unknown but the disease is rare, estimates can be obtained using the \emph{rare disease approximation} that $\pi_1 \approx 0$, which typically introduces negligible bias \citep{Stalder2017}.

\citet{Wang2018unpublished} proposed an improvement to the method of \citet{Stalder2017} that increases the efficiency of the estimates with no additional assumptions.  This development relies on the observation that the method of \citeauthor{Stalder2017} removes dependence on the distribution of the genetic and environmental variables in two different fashions; by treating the genetic and environmental variables symmetrically \citeauthor{Wang2018unpublished} generate two sets of parameter estimates that are combined to generate a more efficient estimate.

\subsection{Implementation}

The semiparametric method of \citet{Stalder2017} is implemented as the function \code{spmle} in \pkg{caseControlGE}, detailed in section \ref{sec:spmle}.  Estimating the semiparametric profile likelihood is a computationally intensive process, and significant effort was invested in speeding up calculations.  Estimation functions, including the analytic gradient and hessian, are written in C++ and compiled using \pkg{Rcpp} \citep{Eddelbuettel2013Rcpp}, providing a tremendous speedup over native \proglang{R} code.  Extensive benchmarking and code profiling was conducted, and estimation functions were written to apply matrix operations to contiguous blocks of memory whenever possible, reducing memory latency and allowing modern processors to exploit data level parallelism and perform the same operation on multiple data points simultaneously.

The estimated semiparametric likelihood is maximized using the quasi-Newton optimizer \pkg{ucminf} \citep{Nielsen2016ucminf} using starting values from logistic regression.  \code{ucminf} is particularly well suited for this application because it allows us to precondition the optimization with the analytic hessian, and it evaluates the gradient after each call to the objective function.  Calculating the gradient along with the likelihood adds negligible computational complexity, so we call a single C++ function to compute them both, then return them separately to \code{ucminf}.  This leads \code{ucminf} to converge in roughly half the time of the next-fastest optimizers (several of the various \proglang{R} implementations of the BFGS algorithm tie for second place).  The unmatched speed of \code{ucminf} means we are willing to tolerate its bugs, which include occasionally declaring convergence before actually converging.  To address this, \code{spmle} checks the gradient at the reported optimum and restarts the optimization if necessary (with different starting values).

Computational complexity of the asymptotic covariance estimation, which contains a sum of the form $\sum_{i=1}^{n} \: \sum_{j=1}^{n} \: \sum_{k=1}^{n} \partial {\cal L}_{ijk}(\Omega) / \partial \Omega$, was reduced from $O(n^3)$ to $O(n^2)$ by storing intermediate values in a three-dimensional array.  This increases speed at the cost of memory usage, which climbs from $O(n)$ to $O(n^2)$, setting a practical limit on sample size in the low tens of thousands for average personal computers.  This is sufficient to analyze all but the largest case-control studies; covariance estimates for larger studies should be computed using the bootstrap.

Asymptotic covariance estimates for the Symmetric Combination Estimator of \citeauthor{Wang2018unpublished} converge slowly and unreliable in practice, often providing poor coverage.  \citeauthor{Wang2018unpublished} recommend a balanced bootstrap, with cases and controls resampled separately, to estimate covariance.  \pkg{caseControlGE} offers users with multicore computers the option to speed up computation by using multiple processors.  Parallelization is implemented using the \proglang{R} base package \pkg{parallel}, which is installed by default on all operating systems.  Parallelization on computers running Linux or macOS is done by forking the active \proglang{R} session, saving time and memory.  This option is unavailable in Windows, so parallelization is fractionally slower because a PSOCK cluster is created with a new instance of \proglang{R} running on each core.


\section{Simulating case-control data with simulateCC} \label{sec:simulateCC}

\subsection{Data description}

\cite{Wang2018unpublished} demonstrate the utility of their method by analyzing data from a case-control study of breast cancer.  This case-control sample is taken from a large prospective cohort at the National Cancer Institute: the Prostate, Lung, Colorectal and Ovarian (PLCO) cancer screening trial \citep{canzian2010comprehensive}.  The case-control study analyzed by \citeauthor{Wang2018unpublished} consists of 658 cases and 753 controls sampled from a cohort of 64,440 non-Hispanic, white women aged 55 to 74, of whom 3.72\% developed breast cancer \citep{pfeiffer2013risk}.  The data are available from the National Cancer Institute via a data transfer agreement, but cannot be distributed with the \pkg{caseControlGE} package.  Fortunately, we can use the \pkg{caseControlGE} function \code{simulateCC} to generate a similar data set for analysis.

Each of the 1411 subjects in the PLCO sample was genotyped for 21 SNPs that have been previously associated with breast cancer based on large genome-wide association studies.  These SNPs were weighted by their log-odds-ratio coefficients and summed to define a polygenic risk score (PRS).  A standardized version of this PRS, with mean zero and standard deviation one, was used as the genetic risk factor $G$ by \citeauthor{Wang2018unpublished}.  Early menarche is a known risk factor for breast cancer, and \citeauthor{Wang2018unpublished} used a binary indicator of whether the subject underwent early menarche as $E$ (age at menarche < 14).  Several environmental variables were recorded as part of the PLCO study, including body mass index (BMI).  In our simulation, we will consider BMI in addition to the variables modeled by \citeauthor{Wang2018unpublished}.

\subsection{Data simulation}

To determine the appropriate distributions to use when simulating $G$ and $E$, we examine the PLCO data.  In doing so, it is important to keep in mind that the case control sample is not representative of the source population.  Case-control studies deliberately oversample cases, so the distribution of $G$ and $E$ in the sample may be quite different from the distribution of $G$ and $E$ in the population (especially for variables that are strongly correlated with disease status).  \code{simulateCC} works by simulating a population using user-specified parameters, then selecting \code{ncase} cases and \code{ncontrol} controls as the case-control sample.  To accurately simulate the genetic and environmental variables from the PLCO study, we need to estimate their distributions \emph{in the source population}.  

The simplest and most common way to estimate population parameters is to calculate them using just the controls.  Case-control designs are typically used to study relatively rare diseases, and the bias introduced by using the cases as a stand-in for the population is usually quite small.  This \emph{rare disease approximation} of population parameters is the best possible estimate when the true population disease rate $\pi_1$ is unknown or not well estimated.  When $\pi_1$ is known, it is possible to calculate unbiased estimates by weighting the cases and controls by $\pi_1$ and $(1 - \pi_1)$, respectively.  (This technique is employed to great effect by \citeauthor{Stalder2017}, and is the reason that \code{spmle} requires the user to specify a value for \code{pi1}.)

\citeauthor{Wang2018unpublished} report $\beta_G = 0.459$ with $p < 1e-4$, but they standardized $G$ to mean zero and standard deviation one \emph{in the case-control sample}.  $G$ has a strong positive effect on disease risk, indicating that the distribution of $(G|D=0)$ is meaningfully different from the distribution of $(G|D=1)$.  Specifically, $\mathbf{E}(G|D=1) > \mathbf{E}(G|D=0)$.  With a population disease rate of 0.0372, this implies $\mathbf{E_{\rm pop}}(G) \approx \mathbf{E}(G|D=0) < 0$, where the subscript pop emphasizes that the expectation is in the source population.  This causes no problem for \citeauthor{Wang2018unpublished}, but it presents us with a dilemma: $G \nsim \hbox{N}(0,1)$.  If we simulate $G \sim \hbox{N}(0,1)$ and use $\beta_G = 0.459$ as reported in \citeauthor{Wang2018unpublished}, we 



should try to approximate the d

To perform a realistic simulation we 

Rather than trying


The entire population is used to calculate disease prevalence (reported
when \code{control$trace > -1}).  



yields the information necessary to simulate the genetic and environmental variables.


- $G$ is a standardized polygenic risk score - simulate from N(0,1)


<!-- ## ncase = 658 -->
<!-- ## ncontrol = 753 -->
<!-- ## pi1 = 2397/64440  = approx 0.037 -->
<!-- ## G is a standardized polygenic risk score - simulate from N(0,1) -->
<!-- ## E1 is a binary indicator of early age at menarche, and has an expected E_bin_freq = 0.745 in the population (weighted expectation by case-control status almost identical to case-only average) -->
<!-- ## E2 is standardized BMI - it is slightly light-tailed, but roughly approximately normal -->





```{r}
1
```


\subsection{Confirming the G-E independence assumption}



\section{Analyzing case-control data with spmle} \label{sec:spmle}


```{r}
1
```

\section{Analyzing case-control data with spmleCombo} \label{sec:spmleCombo}


\bibliography{RJCrefs_05-30-2018}


<!-- \includepdf[pages=-, fitpaper=true]{/home/alex/Research/Tianying/caseControlGE/caseControlGE} -->
